---
title: "Final Project Proposal"
author: "Danilo Canivel, Joel Zou, Danny Breyfogle"
date: "7/15/2019"
output: html_document
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Group Introduction

This study will be conducted by the following students of Stat420 Summer 2019:

 - Danilo Canivel (canivel2)
 - Joel Zou (joelzou2)
 - Danny Breyfogle (dwb4)

## Let's call a Taxi... Maybe not.

## Dataset

For this project we will be predicting the fare amount (inclusive of tolls) for a taxi ride in New York City given the pickup and dropoff locations. 

This dataset has 6 features:

- `pickup_datetime` (datetime)
- `pickup_longitude` (numeric)
- `pickup_latitude` (numeric)
- `dropoff_longitude` (numeric)
- `dropoff_latitude` (numeric)
- `passenger counts` (numeric)

We will be predicting `fare_amount`, fitting on a training set, and predicting on a test set.

There are about 55 million observations in the training set. This is a very large dataset, so we may need to use a subset to meet computing requirements. In order to satisfy the project requirement of at least one categorical predictor, we can transform the `pickup_datetime` feature into various categorical variables such as day of the week, morning, after, evening, etc. that would potentially be helpful in determining the fare amount.

We are going to set a preliminary objective of trying to reach a $RMSE < 2$ but a $RMSE < 2.88$ can be considered acceptable. After predicting, we are going to compare with the Uber estimate API and deliver both results to the customer in a simple Web interface for decision.

The dataset can be found at: [https://www.kaggle.com/c/new-york-city-taxi-fare-prediction/data]()

The Uber API can be found here:
[https://developer.uber.com/docs/riders/references/api/v1.2/estimates-price-get]()

Finally, below are a few observations from the training set:

```{r}
taxi_data = read.csv("input/train.csv", nrows=100000)
head(taxi_data)
```