---
title: "Final Project Proposal"
author: "Danilo Canivel, Joel Zou, Danny Breyfogle"
date: "7/15/2019"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Group Introduction

This study will be conducted by the following students of Stat420 Summer 2019:

 - Danilo Canivel (NetID: canivel2)
 - Joel Zou (NetID: joelzou2)
 - Danny Breyfogle(NetID: dwb4)

## Title

Let's call a Taxi... Maybe not.

## Dataset

For this project we will be predicting the fare amount (inclusive of tolls) for a taxi ride in New York City given the pickup and dropoff locations. We are gonna set a prelimirar objective of trying to reach a RMSE < $2 but a RMSE < 2.88 can be considered acceptable.
After predicting we are going to compare with the Uber estimate API and deliver both results to the customer in a simple Web interface for decision.

The dataset can be found at: https://www.kaggle.com/c/new-york-city-taxi-fare-prediction/data

The Uber API can be found here:
https://developer.uber.com/docs/riders/references/api/v1.2/estimates-price-get

```{r}
taxi_data = read.csv("input/train.csv", nrows=100000)
head(taxi_data)
```


